\section{Introduction}

Weighted Histogram Analysis Method has been widely used in the field of 
molecular dynamics simulations in calculating the potential of mean force
(PMF)~\cite{Kumar1992,Roux1995275}.
It has also been used in free energy calculation in a alchemical scheme~\cite{Kumar1992} as 
well as estimating the ensemble average from generalized ensemble simulations~\cite{Chodera2007}.
However, commonly used implementation of WHAM such as Alan Grossfield's 
implementation~\cite{wham205} and GROMACS ``g\_wham" only support PMF calculation.
Moreover, most implementations uses the direct iteration to solve for PMFs 
while it's been shown~\cite{Zhu2012} that there are more efficient way to 
reach the same answers. Here I combine the reasonings behind the aforementioned
literature and put together an implementation of WHAM that's capable of solving 
the following general problem:

Suppose the we have $K$ systems of the same thermodynamic ensemble with
each characterized by different thermodynamics parameters. If we let $\vect{x} \in
 \nspace{R}{N}$ being the generalized coordinates, which has $N$  degrees of 
 freedom, and
\begin{equation}\label{eq:Ham}
H_{i}(\vect{x}) = \beta_{i}[ U_{i}(\vect{x}) + P_{i}V(\vect{x}) + \dotprd{\boldsymbol\mu_{i}}{\vect{n}(\vect{x})}]
\end{equation}
be the Hamiltonian of the $i$'th system, where $\beta_{i}, U_{i}, P_{i}, V, \boldsymbol\mu_{i}$
and $\vect{n}$ are the inverse temperature, potential energy, the pressure, the volume, the chemical
potentials for each system components and the number of molecules for each components, 
respectively. The potential energy $U_{i}$ can be further decomposed into 
the internal potential $U$ and $L$ external potentials $\vect{U}_{b}$, both 
of which are of the same form across the $K$ systems, so that 
\begin{equation}\label{eq:Pot}
U_{i}(\vect{x}) = U(\vect{x}) + \dotprd{\boldsymbol\lambda_{i}}{\vect{U}_{b}}
\end{equation} 
Each of the $K$ systems can has its own specific $\boldsymbol\lambda_{i}$,  
which is often used in the alchemical calculations. Here both $\boldsymbol\lambda_{i}$ 
and $\vect{U}_{b}$ are of the dimension of $L$. To simplify the expression, 
we rewrite equation~\ref{eq:Ham} into
\begin{equation}\label{eq:simHam}
H_{i}(\vect{x}) = \beta_{i}\dotprd{\vect{R}_{i}}{\vect{q}(\vect{x})}
\end{equation}
where 
\begin{equation}
\vect{R}_{i} \equiv \{1, \boldsymbol\lambda_{i}, P_{i}, \boldsymbol\mu_{i}\}
\label{eq:R}
\end{equation}
and
\begin{equation}
\vect{q}(\vect{x}) \equiv \{U(\vect{x}), \vect{U}_{b}(\vect{x}), V(\vect{x}), \vect{n}(\vect{x})\}
\label{eq:q}
\end{equation}
are the arrays of intensive and extensive parameters, respectively. 
We further define unnormalized probability density 
\begin{equation}\label{eq:p}
p_{i}(\vect{x}) \equiv \exp(-H_{i}(\vect{x}))
\end{equation}
and the partition function
\begin{equation}\label{eq:Q}
Q_{i} \equiv \intgrln{}{}{p_{i}}{x} 
\end{equation}

Now we'd like to estimate the ensemble average of some observable 
$O \equiv O(\vect{x})$ in any of the system $i$
\begin{equation}\label{eq:O}
\eavg{O}_{i} \equiv \frac{\intgrln{}{}{p_{i}(\vect{x})O}{x}}{Q_{i}}
\end{equation}
For example, the PMF of a reaction coordinate 
$\boldsymbol\xi \equiv \boldsymbol\xi(\vect{x})$ is
\begin{equation}
\label{eq:PMF}
w(\boldsymbol\xi) \equiv \eavg{\delta(\boldsymbol\xi - \boldsymbol\xi(\vect{x}))} 
\end{equation}
where $\delta$ is the Dirac delta function.
Or in other cases we want to know the free energy difference between 
any two systems $i$ and $j$:
\begin{equation}
\dG_{ij} \equiv -k_{B}T \ln\frac{Q_{j}}{Q_{i}}
\label{eq:dG}
\end{equation}

To do that, we carry out $S$ simulations ($S \ge K$) with each 
simulation visiting at least $1$ of the $K$ systems. As soon as the transition 
from one systems to the other follows the same statistics as those used 
in transition between configuration in any of the systems, we can simply 
assert $S = K$ for simplicity and the cases where $S > K$ can be handled 
accordingly without troubles at all. The purpose of WHAM is to provide 
an optimal estimate of $O$ and $\dG_{ij}$ by combining the estimates from 
all the $S$ simulations. 
